---
title: "Solution for Network Analysis"
author: "Florian Klaver"
date: "2026-04-18"
format: html
---

In this document, I solve the tasks for *network analysis* of the course *Spatiotemporal Datascience*.

## Task 1: Centrality
In this task, you will analyze the cycle network of Switzerland using R. The dataset comes from the [Federal Roads Office FEDRO](https://data.geo.admin.ch/browser/index.html#/collections/ch.astra.veloland) (ASTRA).

The goal is to compute different centrality measures (degree, betweenness and closeness) for the nodes in the network and visualize the results. To do this, follow the following steps:

1. Load the rail network data using `sf`
2. Convert the data into an *undirected* network using `sfnetworks`
3. To calculate centrality, the edges need to have "weights". Assuming equal cycling speed on all roads, you can use the edge length (from the column *length*) as the weight. 
4. To compute the centrality measures you will need to 
   1. activate the *nodes* (using `tidygraph::activate()`) 
   2. Use `dplyr::mutate` to create new columns and populate these with the centrality measures using the functions `centrality_betweenness`, `centrality_closeness` and `centrality_degree` from `tidygraph` using the column *length* as weights.
5. Convert the enriched network consisting of edges and nodes back to `sf` objects consisting of lines and points (using `activate` and `st_as_sf`)
6. The centrality values are attached to the points. Transfer the values to the lines by using the following function, which takes the mean centrality values over all nodes for each edge
  
   ```{.r}
   aggregate(veloland_nodes, veloland_edges, FUN = "mean")
   ```
7. Visualize the result. If necessary, transform or reasonably categorize the values to make them more comparable
8. Discuss the results: How do you interpret the differences?

### Preparation
```{r}
#| warning: FALSE
#| message: FALSE

# Set working directory
setwd("C:/Workspace/ZHAW/6_Semester/STDS/network-analysis")

# Load libraries
library(sf)
library(sfnetworks)
library(tidygraph)
library(dplyr)
library(ggplot2)
library(patchwork)

```

### Analysis
```{r}
# Check available layers in the database
st_layers("data/veloland.gpkg")

##  Step 1: Load the Veloland dataset
velo <- st_read("data/veloland.gpkg")
colnames(velo)

# Check for duplicates and NAs
sum(duplicated(velo))
sum(is.na(velo))


## Step 2: Convert to an undirected network
velo_net <- as_sfnetwork(velo, directed = FALSE)

## Step 3 and 4: Calculate centrality measures
# Activate nodes and use SHAPE_Length as the cost/weight for traveling between nodes
velo_net_centrality <- velo_net |>
  activate("nodes") |>
  mutate(
    degree = centrality_degree(weights = length),
    betweenness = centrality_betweenness(weights = length),
    closeness = centrality_closeness(weights = length)
  )


## Step 5: Convert back to sf objects
# Extract nodes as points, select only the calculated numeric columns
veloland_nodes <- velo_net_centrality |>
  activate("nodes") |>
  st_as_sf() |>
  select(degree, betweenness, closeness)

# Extract edges as lines
veloland_edges <- velo_net_centrality |>
  activate("edges") |>
  st_as_sf()


## Step 6: Transfer node values to lines
veloland_edges_enriched <- aggregate(
  x = veloland_nodes, 
  by = veloland_edges, 
  FUN = mean,
  na.rm = TRUE # handle potential NA values
)

# Add log1p transforms for skewed measures to improve color variation in maps
veloland_edges_enriched <- veloland_edges_enriched |>
  mutate(
    betweenness_log1p = log1p(betweenness),
    degree_log1p = log1p(degree)
  )

# Quick diagnostics to spot problematic centrality outputs
if (any(!is.finite(veloland_nodes$closeness), na.rm = TRUE)) {
  warning("Non-finite closeness values detected. The network may contain disconnected components.")
}

centrality_summary <- veloland_nodes |>
  st_drop_geometry() |>
  summarise(
    across(
      c(degree, betweenness, closeness),
      list(
        min = ~ min(.x, na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        max = ~ max(.x, na.rm = TRUE),
        na = ~ sum(is.na(.x)),
        inf = ~ sum(is.infinite(.x))
      )
    )
  )

centrality_summary

```

### Visualization
```{r}
#| fig-width: 8
#| fig-height: 14
## Define common theme for all maps
map_theme <- theme_void() +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 11, margin = margin(t = 6, b = 4)),
    legend.position = "bottom",
    legend.key.width = unit(2.0, "cm"),
    legend.key.height = unit(0.2, "cm"),
    legend.title = element_text(vjust = 0.85, size = 8.5),
    legend.text = element_text(size = 8),
    legend.box.margin = margin(t = -2, b = 4),
    plot.margin = margin(t = 8, r = 8, b = 10, l = 8)
  )

## Individual plots
# Degree Centrality
p_degree <- ggplot(data = veloland_edges_enriched) +
  geom_sf(aes(color = degree_log1p, linewidth = degree_log1p), lineend = "round") +
  scale_color_viridis_c(option = "viridis") +
  scale_linewidth_continuous(range = c(0.2, 1.0), guide = "none") +
  labs(
    title = "Degree Centrality",
    color = "log1p(Degree)"
  ) +
  coord_sf(datum = NA) +
  map_theme

# Closeness Centrality
p_closeness <- ggplot(data = veloland_edges_enriched) +
  geom_sf(aes(color = closeness, linewidth = closeness), lineend = "round") +
  scale_color_viridis_c(option = "viridis") +
  scale_linewidth_continuous(range = c(0.2, 1.0), guide = "none") +
  labs(
    title = "Closeness Centrality",
    color = "Closeness"
  ) +
  coord_sf(datum = NA) +
  map_theme

# Betweenness Centrality
p_betweenness <- ggplot(data = veloland_edges_enriched) +
  geom_sf(aes(color = betweenness_log1p, linewidth = betweenness_log1p), lineend = "round") +
  scale_color_viridis_c(option = "viridis") +
  scale_linewidth_continuous(range = c(0.2, 1.0), guide = "none") +
  labs(
    title = "Betweenness Centrality",
    color = "log1p(Betweenness)"
  ) +
  coord_sf(datum = NA) +
  map_theme

## Combine and render
combined_plot <- (p_degree / p_closeness / p_betweenness) +
  plot_annotation(
    title = "Swiss National Cycle Network: Edge Centrality",
    caption = "Data: ASTRA FEDRO | Edge-level mean aggregation",
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold", margin = margin(t = 10, b = 10)),
      plot.caption = element_text(size = 9, margin = margin(t = 12, b = 4))
    )
  )

## Print the final result
combined_plot
```

### Discussion
The centrality analysis of the Swiss cycle network reveals spatial patterns, some more clearer than others. Degree Centrality seems very similar for many of the edges. There is not much variation and I can't see any distinct patterns. The road sticking out the most is the "ring" in the north going around what i assume to be Basel. It has the highest degree centrality and must have many connections. Closeness centrality is highest on the Swiss Plateau, which sits more or less in the geographic center of the country and and holds a large part of the population. It is also visible that there the nodes more dense than for example in the southern regions where the alps are. So both geographically and population density wise this can be expected and makes absolutely sense. Betweenness highlights a handful of corridors, mainly the main east-west routes that act as bridges for the whole network (particularly the one bright yellow road running through the Swiss Plateau). Similar to the degree map, betweenness has low variation with only a few single roads sticking out.

One clear limitation is that we used physical edge length as the only weight and therefore factors such as steepness are not considered. Including elevation data (e.g., a digital elevation model) to estimate actual travel effort or time would give more detailed results. Another alternative would be to use actual cycling speed or infrastructure quality as weights.

## Task 2: Routing
In this task, you will do your do your own routing using the Veloland dataset from the previous task (locations.gpkg on moodle). Import the two datasets and convert the Veloland dataset to an undirected network using `sfnetworks`, as you did before.

1. Subset a single location and call this `origin`
2. Create a second object with all the other locations and call this `destination`
3. Use the function `st_network_paths` from `sfnetworks` to calculate the shortest paths from `origin`, to `destination`, using the column `length` as weight
4. The output returns a two column data.frame. The column `edge_paths` retuns the row numbers of the line segments that constitute the shortest path. Use these values to create an `sf` object for each path.
5. Visualize the result in a map. Include `origin` and `destination` in your visualization.

### Preparation
```{r}
st_layers("data/locations.gpkg")

locations <- st_read("data/locations.gpkg")

# Make sure the location points use the same CRS as the Veloland network from Task 1.
if (st_crs(locations) != st_crs(veloland_edges)) {
  locations <- st_transform(locations, st_crs(veloland_edges))
}
```

### Analysis
```{r}

# Reuse the Veloland network and edge geometries from Task 1.
# Define origin and destination points
# I choose the first location as origin and the rest as destinations
origin <- locations |> slice(1)
destination <- locations |> slice(-1) |> mutate(destination_id = row_number())

# Calculate shortest paths from origin to every destination using edge length as weight.
route_paths <- st_network_paths(
  x = velo_net,
  from = origin,
  to = destination,
  weights = "length"
)

# Convert each edge path into a single sf geometry for plotting.
route_geometries <- lapply(route_paths$edge_paths, function(edge_ids) {
  edge_ids <- edge_ids[!is.na(edge_ids)]

  if (length(edge_ids) == 0) {
    return(st_geometrycollection())
  }

  st_line_merge(st_union(st_geometry(veloland_edges[edge_ids, ])))[[1]]
})

route_results <- route_paths |>
  st_drop_geometry() |>
  mutate(
    destination_id = destination$destination_id,
    n_edges = lengths(edge_paths)
  )

route_sf <- st_sf(
  route_results,
  geometry = st_sfc(route_geometries, crs = st_crs(veloland_edges))
)

route_sf$route_color <- grDevices::hcl.colors(nrow(route_sf), palette = "Dark 3")

```

### Visualization
```{r}

ggplot() +
  geom_sf(data = veloland_edges, color = "grey90", linewidth = 0.12) +
  geom_sf(aes(color = route_color), data = route_sf, linewidth = 0.6, alpha = 0.9, show.legend = FALSE) +
  geom_sf(aes(fill = "Destination"), data = destination, shape = 21, color = "white", size = 2.6, alpha = 0.9) +
  geom_sf(aes(fill = "Origin"), data = origin, shape = 21, color = "white", size = 3.4) +
  labs(
    title = "Shortest Paths from One Origin to All Other Locations",
    fill = NULL,
    caption = "Background: Veloland network | Routes computed with length-weighted shortest paths"
  ) +
  scale_color_identity() +
  scale_fill_manual(values = c(Origin = "#1B9E77", Destination = "#2C7FB8")) +
  coord_sf(datum = NA) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0.5),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  )


```

### Discussion
The routing analysis successfully computes the shortest paths from a single origin to multiple destinations across the Swiss cycle network. The visualization clearly shows the main corridors connecting the origin to various points, with some routes sharing common segments, especially those heading towards the central and eastern parts of Switzerland. The use of edge length as weights ensures that the computed paths reflect actual travel distances, which is crucial for realistic routing. However, similar to the centrality analysis, this approach does not account for factors such as elevation changes or other factors. So for this sole purpose the analysis works well, but for more detailed routing recommendations, incorporating additional data (e.g., elevation, traffic conditions, or infrastructure quality) would improve the accuracy and relevance of the routes generated.

## Use of AI
I tried to solve the tasks mostly without the help of AI, but I used Gemini correct my solutions and improve some of my code if I was stuck or unsure.